---
title: "Jiexin-Yan_2026 scATAC-seq of LSK cells from naive or DM mice"
author: "WuYuqian"
date: "2026-03-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

scATAC-seq of LSK cells from naive or DM mice.

## Packages

```{r}
library(Signac)
library(Seurat)
library(ArchR)
library(dplyr)
library(GenomicRanges)
library(future)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(AnnotationHub)
library(rtracklayer)
library(monocle3)
library(presto)
library(JASPAR2020)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10)
library(SeuratWrappers)
plan("multicore", workers = 8)
options(future.globals.maxSize = 200 * 1024^3)
```

## Data loading

Following <https://stuartlab.org/signac/articles/overview>

```{r}
# Peaks and annotations
peaks.CON <- read.table(
  file = "/home/YJX-scATAC/CON-processed/outs/peaks.bed",
  col.names = c("chr", "start", "end")
)
peaks.DM <- read.table(
  file = "/home/YJX-scATAC/DM-processed/outs/peaks.bed",
  col.names = c("chr", "start", "end")
)

gr.CON <- makeGRangesFromDataFrame(peaks.CON)
gr.DM <- makeGRangesFromDataFrame(peaks.DM)
combined.peaks <- reduce(x = c(gr.CON, gr.DM))
peakwidths <- width(combined.peaks)
combined.peaks <- combined.peaks[peakwidths  < 10000 & peakwidths > 20]
ah <- AnnotationHub()
query(ah, "EnsDb.Mmusculus.v98")
ensdb_v98 <- ah[["AH75036"]]
annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "mm10"
```

Fragment matrix

```{r}
md.CON <- read.table(
  file = "/home/YJX-scATAC/CON-processed/outs/singlecell.csv",
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ] # remove the first row
md.DM <- read.table(
  file = "/home/YJX-scATAC/DM-processed/outs/singlecell.csv",
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]
md.CON <- md.CON[md.CON$passed_filters > 1000, ]
md.DM <- md.DM[md.DM$passed_filters > 1000, ]
frags.CON <- CreateFragmentObject(
  path = "/home/YJX-scATAC/CON-processed/outs/fragments.tsv.gz",
  cells = rownames(md.CON)
)
frags.DM <- CreateFragmentObject(
  path = "/home/YJX-scATAC/DM-processed/outs/fragments.tsv.gz",
  cells = rownames(md.DM)
)
CON.counts <- FeatureMatrix(
  fragments = frags.CON,
  features = combined.peaks,
  cells = rownames(md.CON)
)
DM.counts <- FeatureMatrix(
  fragments = frags.DM,
  features = combined.peaks,
  cells = rownames(md.DM)
)
```

Create Chromatin Assay and annotation

```{r}
CON_assay <- CreateChromatinAssay(CON.counts, fragments = frags.CON)
CON <- CreateSeuratObject(CON_assay, assay = "ATAC", meta.data=md.CON)
DM_assay <- CreateChromatinAssay(DM.counts, fragments = frags.DM)
DM <- CreateSeuratObject(DM_assay, assay = "ATAC", meta.data=md.DM)
CON$dataset <- 'CON'
DM$dataset <- 'DM'
Annotation(CON) <- annotations
Annotation(DM) <- annotations
```

## Quality control

```{r}
CONpeaks.keep <- seqnames(granges(CON)) %in% standardChromosomes(granges(CON))
CON <- CON[as.vector(CONpeaks.keep), ]
DMpeaks.keep <- seqnames(granges(DM)) %in% standardChromosomes(granges(DM))
DM <- DM[as.vector(DMpeaks.keep), ]
CON <- NucleosomeSignal(object = CON)
DM <- NucleosomeSignal(object = DM)
CON <- TSSEnrichment(object = CON)
DM <- TSSEnrichment(object = DM)
CON$pct_reads_in_peaks <- CON$peak_region_fragments / CON$passed_filters * 100
DM$pct_reads_in_peaks <- DM$peak_region_fragments / DM$passed_filters * 100
mm10_blacklist <- import("mm10.blacklist.bed")
CON$blacklist_ratio <- FractionCountsInRegion(
  object = CON, 
  assay = 'ATAC',
  regions = mm10_blacklist
)
DM$blacklist_ratio <- FractionCountsInRegion(
  object = DM, 
  assay = 'ATAC',
  regions = mm10_blacklist
)

CON <- subset(
  x = CON,
  subset = peak_region_fragments > 5000 &
    peak_region_fragments < 100000 &
    pct_reads_in_peaks > 40 &
    blacklist_ratio < 0.02 &
    nucleosome_signal < 2 &
    TSS.enrichment > 2
)

DM <- subset(
  x = DM,
  subset = peak_region_fragments > 5000 &
    peak_region_fragments < 100000 &
    pct_reads_in_peaks > 40 &
    blacklist_ratio < 0.02 &
    nucleosome_signal < 2 &
    TSS.enrichment > 2
)
```

## Dataset integration

```{r}
combined <- merge(
  x = CON,
  y = DM,
  add.cell.ids = c("CON", "DM")
)
```

## Dimensionality reduction and clustering

```{r}
set.seed(1234)
combined <- RunTFIDF(combined)
combined <- FindTopFeatures(combined, min.cutoff = 'q5')
combined <- RunSVD(combined)
combined <- RunUMAP(combined, dims = 2:30, reduction = 'lsi')
combined <- FindNeighbors(object = combined, reduction = 'lsi', dims = 2:30)
combined <- FindClusters(object = combined, algorithm = 3) 
```

## Gene activity matrix

```{r}
gene.activities <- GeneActivity(combined)
combined[['RNA']] <- CreateAssayObject(counts = gene.activities)
combined <- NormalizeData(
  object = combined,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(combined$nCount_RNA)
)
```

## Differential accessibility analysis

Using gene activity matrix to perform DAR analysis

```{r}
DefaultAssay(combined) <- "RNA"
Idents(combined) <- "dataset"
DAR <- FindMarkers(
  combined,
  ident.1 = "DM",
  ident.2 = "CON",
  assay = "RNA",
  only.pos = FALSE,
  test.use = "LR", 
  latent.vars = "nCount_ATAC",
  logfc.threshold = 0,
  min.pct = 0.05
)
```

## Cell Percentage

```{r}
Idents(combined) <- "seurat_clusters"
cluster_table <- table(Idents(combined), combined$dataset)
total_cells_per_group <- colSums(cluster_table)
cluster_fraction <- sweep(cluster_table, 2, total_cells_per_group, FUN = "/")
cluster_percent <- apply(cluster_fraction, 1, function(x) {
  total <- sum(x)
  round(100 * x / total, 2)
})
cluster_percent <- t(cluster_percent)
write.csv(cluster_percent, file = "cluster_CON_DM_percentage.csv")
df <- read.csv("cluster_CON_DM_percentage.csv", row.names = 1)
colnames(df) <- c("Ctrl", "DM")
df$cluster <- as.integer(rownames(df)) + 1
df_long <- melt(df, id.vars = "cluster", variable.name = "group", value.name = "percent")
df_long$cluster <- factor(df_long$cluster, levels = 15:1)
ggplot(df_long, aes(x = group, y = cluster, fill = percent)) +
  geom_text(aes(label = sprintf("%.2f", percent)), size = 3)
  labs(x = "", y = "Cluster", fill = "%") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

## Motif analysis

```{r}
Idents(combined) <- "dataset"
table(Idents(combined))
DefaultAssay(combined) <- "ATAC"  

pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)

# add motif information
combined <- AddMotifs(
  object = combined,
  genome = BSgenome.Mmusculus.UCSC.mm10,
  pfm = pfm
)

da_peaks <- FindMarkers(
  object = combined,
  ident.1 = 'DM',
  ident.2 = 'CON',
  only.pos = TRUE,
  test.use = 'LR',
  min.pct = 0.05,
  latent.vars = 'nCount_ATAC'
)
top.da.peak <- rownames(da_peaks[da_peaks$p_val < 0.005 & da_peaks$pct.1 > 0.2, ])
enriched.motifs <- FindMotifs(
  object = combined,
  features = top.da.peak
)
write.csv(enriched.motifs, file = "enriched_motifs.csv", row.names = FALSE)

motif_df <- read_csv("enriched_motifs.csv")
motif_df <- motif_df %>%
  mutate(logP = -log10(p.adjust))#%>%

highlight_list <- list(
  "Positive regulation of glycolytic process" = c("MLX", "Arnt", "MLXIPL", "MYC"),
  "Macrophage differentiation" = c("CEBPE", "CEBPA", "SPIB", "FOSL2", "SPI1"),
  "Myeloid cell differentiation" = c(
    "FOS", "BATF", "CREB1", "CEBPE", "JUN", "SP1", "CEBPA", "JUNB", 
    "CEBPB", "RUNX1", "KLF2", "SPIB", "SP3", "SPI1", "BATF3", "CEBPG", "ZBTB7A"
  )
)

for (go_term in names(highlight_list)) {
  highlight_motifs <- highlight_list[[go_term]]
  motif_df$highlight <- motif_df$motif.name %in% highlight_motifs
  file_name <- paste0("motif_volcano-", go_term, ".pdf")
  pdf(file_name, width = 5, height = 5)
  print(
    ggplot(motif_df, aes(x = fold.enrichment, y = logP)) +
      geom_point(color = "grey", alpha = 0.7, size = 1.5) +
      geom_point(data = subset(motif_df, highlight), color = "#a7399b", size = 1.5) +
      geom_text_repel(data = subset(motif_df, highlight),
                      aes(label = motif.name),
                      size = 7,
                      max.overlaps = 100,
                      box.padding = 0.3,
                      point.padding = 0.2) +
      theme_minimal() +
      theme(
        panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.ticks.length = unit(0.25, "cm"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.8),
        legend.title = element_blank(),
        axis.title = element_text(size = 18),
        axis.text.x = element_text(size = 18, vjust = 0),
        axis.text.y = element_text(size = 18, angle = 90, vjust = 0.5, hjust = 0.5),
        plot.title = element_text(hjust = 0.5, size = 18)
      ) +
      labs(
        title = go_term,
        x = "Fold enrichment",
        y = expression(-log[10]*"(p-adj)")
      ) +
      scale_y_continuous(breaks = seq(0, max(motif_df$logP, na.rm = TRUE), by = 5)) +
      scale_x_continuous(breaks = seq(
        floor(min(motif_df$fold.enrichment, na.rm = TRUE)),
        ceiling(max(motif_df$fold.enrichment, na.rm = TRUE)),
        by = 0.5
      ))
  )
  dev.off()
}

```

## VlnPlot and CoveragePlot

```{r}
VlnPlot(
  combined,
  features = "Sirpa",
  assay = "RNA",
  slot = "data",
  pt.size = 0
) + NoLegend() +
  scale_fill_manual(values = color_codes)

CoveragePlot(
  object = combined,
  region = "Sirpa",
  annotation = TRUE,
  peaks = TRUE,
  group.by = "dataset",
  window = 5,
  extend.upstream = 5000,
  extend.downstream = 5000
)
```

## Pseudotime trajectory analysis

### Overall plot

```{r}
DefaultAssay(combined) <- "ATAC"
cds_rna <- as.cell_data_set(combined)
cds_rna <- cluster_cells(cds_rna, reduction_method = "UMAP")
cds_rna <- learn_graph(cds_rna, 
                       use_partition = TRUE, 
                       learn_graph_control = list( 
                         euclidean_distance_ratio=0.1))  
get_root_principal_node_by_cluster <- function(cds_rna, cluster_id = "9") {
  cell_ids <- which(colData(cds_rna)$seurat_clusters == cluster_id)
  closest_vertex <- cds_rna@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds_rna), ])
  root_pr_node <- igraph::V(principal_graph(cds_rna)[["UMAP"]])$name[
    as.numeric(names(which.max(table(closest_vertex[cell_ids, ]))))
  ]
  return(root_pr_node)
}
cds1 <- order_cells(cds_rna, 
                    root_pr_nodes = get_root_principal_node_by_cluster(cds_rna, cluster_id = "9"),reduction_method = "UMAP")
plot_cells(cds1,
           color_cells_by = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           graph_label_size = 1.5,
           #cell_size = 0.1, 
           trajectory_graph_color = "black")+ coord_equal()
```

### Plotting selected region

```{r}
cds_sub <- choose_graph_segments(cds1, reduction_method = "UMAP")
cells_segment <- colnames(cds_sub)
colData(cds1)$highlight <- "Other"
colData(cds1)$highlight[colnames(cds1) %in% cells_segment] <- "Selected"

plot_cells(
  cds1,
  color_cells_by = "highlight",
  label_cell_groups = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE,
  label_roots = FALSE,
  graph_label_size = 1.5,
  trajectory_graph_color = "black"
) + scale_color_manual(values = c("Other" = "lightgrey", "Selected" = "#a020f0")) + coord_equal()+
  guides(color = guide_legend(override.aes = list(shape = 16))) +
  theme(legend.position = "none")
```

### ArchR-style plotting

Choose selected cells to plot gene accessibility along the trajectory

```{r}
cds_sub <- choose_graph_segments(cds1, reduction_method = "UMAP")
plot_cells(cds_sub,
           color_cells_by = "pseudotime") + coord_equal()
cells_segment <- colnames(cds_sub)
cds_subset <- cds1[, colnames(cds1) %in% cells_segment]
pseudotime_df <- data.frame(
  cell = colnames(cds_subset),
  pseudotime = pseudotime(cds_subset)
)
pseudotime_df$cell <- gsub("^CON_", "Ctrl#", pseudotime_df$cell)
pseudotime_df$cell <- gsub("^DM_", "DM#", pseudotime_df$cell)
write.table(pseudotime_df,
            file = "pseudotime_metadata_sub.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
cells_con <- gsub("^CON_", "", grep("^CON_", cells_segment, value = TRUE))
cells_dm <- gsub("^DM_", "", grep("^DM_", cells_segment, value = TRUE))
writeLines(cells_con, "cells_con_sub.txt")
writeLines(cells_dm, "cells_dm_sub.txt")
```

filter out fragments according to barcodes

```{bash}
zcat /home/YJX-scATAC/CON-processed/outs/fragments.tsv.gz | grep -F -f /home/YJX-scATAC/ArchR/sub_traj/cells_con_sub.txt | bgzip > subset_fragments_CON.tsv.gz
tabix -p bed subset_fragments_CON.tsv.gz

zcat /home/YJX-scATAC/DM-processed/outs/fragments.tsv.gz | grep -F -f /home/YJX-scATAC/ArchR/sub_traj/cells_dm_sub.txt | bgzip > subset_fragments_dm.tsv.gz
tabix -p bed subset_fragments_dm.tsv.gz
```

Create ArchR project (Following <https://www.archrproject.com/bookdown/index.html>)

```{r}
addArchRGenome("mm10")
inputFiles <- c(
  CON = "subset_fragments_CON.tsv.gz",
  EXP = "subset_fragments_dm.tsv.gz"
)
names(inputFiles) <- c("Ctrl","DM")
ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  minTSS = 0,
  minFrags = 0,
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)
proj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = "ArchR2Signac_featureplot_Project",
  copyArrows = TRUE
)
proj <- addIterativeLSI(
  ArchRProj = proj,
  useMatrix = "TileMatrix", 
  name = "IterativeLSI", 
  iterations = 2, 
  clusterParams = list( #See Seurat::FindClusters
    resolution = c(0.2), 
    sampleCells = 10000, 
    n.start = 10
  ), 
  varFeatures = 25000, 
  dimsToUse = 1:30
)
proj <- addClusters(
  input = proj,
  reducedDims = "IterativeLSI",
  method = "Seurat",
  name = "Clusters",
  resolution = 0.8,
  dimsToUse = 2:30
)
proj <- addUMAP(
  ArchRProj = proj, 
  reducedDims = "IterativeLSI", 
  name = "UMAP", 
  nNeighbors = 30, 
  minDist = 0.5, 
  metric = "cosine"
)
```

Plotting gene accessibility along the trajectory

```{r}
pseudo <- read.delim("pseudotime_metadata_sub.tsv",
                     header = TRUE,
                     stringsAsFactors = FALSE)
head(pseudo)
pseudo <- pseudo %>%
  mutate(PseudoCluster = cut_number(pseudotime, n = 7, labels = c(1,2,3,4,5,6,7)))

proj <- addCellColData(
  ArchRProj = proj,
  data = as.character(pseudo$PseudoCluster),
  name = "PseudoCluster",
  cells = pseudo$cell,
  force = TRUE
)
proj <- addTrajectory(
  ArchRProj = proj,
  name = "PseudoCluster_basedTraj",
  trajectory = c("1","2","3","4","5","6","7"),
  groupBy = "PseudoCluster",
  embedding = "UMAP",
  force = TRUE
)
proj <- addImputeWeights(proj)
geneName <- c("Sirpa","Csf1r","Cd68","Itgam")
p1 <- plotTrajectory(
  proj, 
  trajectory = "PseudoCluster_basedTraj", 
  colorBy = "GeneScoreMatrix", 
  name = geneName, 
  continuousSet = "horizonExtra"
)
```

Plotting umap using ImputeWeights in ArchR

```{r}
# extract umap coordinates
umap_signac <- Embeddings(combined, "umap")
head(umap_signac)

signac_cells <- rownames(umap_signac)
archr_cells_map <- gsub("^CON_", "Ctrl#", signac_cells)
archr_cells_map <- gsub("^DM_",  "DM#",   archr_cells_map)

map_df <- data.frame(signac = signac_cells, archr = archr_cells_map, stringsAsFactors = FALSE)
archr_all <- ArchR::getCellNames(proj)
common_archr <- intersect(map_df$archr, archr_all)
rownames(umap_signac) <- archr_cells_map
umap_signac <- as.matrix(umap_signac)
colnames(umap_signac) <- c(
  "IterativeLSI#UMAP_Dimension_1",
  "IterativeLSI#UMAP_Dimension_2"
)
proj@embeddings$SignacUMAP <- SimpleList(
  embedding = umap_signac
)
proj <- addImputeWeights(proj)
geneName <- "Sirpa"
plotEmbedding(
  ArchRProj = proj,
  embedding = "SignacUMAP",   
  colorBy = "GeneScoreMatrix",
  name = geneName,
  ratioYX = 2,
  pal = cont_pal(256)
)
```

# End

```{r}
sessionInfo()
```
